\input{ppreamble}

\newcommand{\tac}[1]{\lstinline[mathescape]~#1~}
\newcommand{\ciff}{\ \leftrightarrow\ }
\newcommand{\hyp}{\tac{H}}
\newcommand{\hypB}{\tac{G}}
\newcommand{\var}{\tac{x}}
\newcommand{\varB}{\tac{y}}

\newtheorem*{nlemma}{Lemma}


\title{Divisibility of multinomial coefficients}
\author{Jules Jacobs}

\begin{document}
\maketitle
\begin{abstract}
  Together with Ike Mulder we discovered a fun little divisibility property of multinomial coefficients (probably well-known!).
\end{abstract}

Our starting point is that not only the binomial coefficient $\frac{(a+b)!}{a! b!}$ is a whole number,
but also the Catalan numbers $\frac{(2n)!}{n! (n+1)!} = \frac{(2n)!}{n!n!} / (n+1)$ are whole numbers.
That $(a+b)!$ is divisible by $a! b!$ is already a small miracle, but the Catalan numbers show that it can be divided even further in certain cases. Our question is: does this generalize to other binomial coefficients?

The answer turns out to be yes: if $\gcd(a, b) = 1$ then $\frac {(a + b - 1)!}{a! b!}$ is a whole number.
Note that this implies the Catalan divisibility because $\gcd(n,n+1) = 1$.

The divisibility generalizes to multinomial coefficients:

\begin{nlemma}
  If $\gcd(a_1, \dots, a_n) = 1$ then
  \begin{align}
    \label{omultinom}
    \frac
      {(a_1 + \cdots + a_n - 1)!}
      {a_1! \cdots a_n!}
  \end{align}
  is a whole number.
\end{nlemma}
\begin{proof}
  From the $\gcd$ assumption, we have integers $k_1,\dots,k_n$ such that
  \begin{align*}
    1 = a_1 k_1 + \cdots + a_n k_n
  \end{align*}
  Multiply both sides by (\ref{omultinom}):
  \begin{align*}
    \frac
      {(a_1 + \cdots + a_n - 1)!}
      {a_1! \cdots a_n!} &=
        \frac
          {(a_1 + \cdots + a_n - 1)!}
          {a_1! \cdots a_n!} a_1 k_1
        + \cdots +
        \frac
          {(a_1 + \cdots + a_n - 1)!}
          {a_1! \cdots a_n!} a_n k_n \\
      &=
      \frac
          {(a_1 + \cdots + a_n - 1)!}
          {(a_1 - 1)! \cdots a_n!} k_1
        + \cdots +
        \frac
          {(a_1 + \cdots + a_n - 1)!}
          {a_1! \cdots (a_n - 1)!} k_n
  \end{align*}
  The right hand side is an integer because multinomial coefficients are integers.
\end{proof}

Slightly more generally, the proof shows that
  $\frac
    {(a_1 + \cdots + a_n - 1)! \gcd(a_1, \dots, a_n)}
    {a_1! \cdots a_n!}$
is always a whole number.
\end{document}